Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies

### 2019-05-16 man/ms.Rmd
update

* The studied relict populations showed expressions of management, likely reflecting the origin of the current forest structure. Both resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and also to long-term climatic changes (*i.e.* warming), despite the ‘a priori’ vulnerability that we were expecting them to express in respone climate change on relict stands resulting from intensive land-use in history at their current rear-edge. Furthermore, the coldest stands show a positive growth trend since the end of the 1970s matching the simultaneous increase in temperature observed.

* Resilience indices showed ecological coherence along a gradient of drought severity in time. In addition, oak resilience to drought followed a dryness gradient among the three populations studied: Northern populations thriving under a drier climate were less resilient to drought both in primary and secondary growth than Southern populations where lower altitudes were also less resilient than those at high-altitudes.

* Despite being considered relict stands at the dry rear-edge for the species, they showed high resilience in response to drought. Yet, the high sensitivity to water availability expressed and positive growth trends only in the coldest and more humid stands suggest divergent response to a changing climate along very short climatic gradients and suggest that dynamics (i.e. not only adult performance but also regeneration) of those stands located at the lowest elevations, hence the altitudinal rear-edge, need to be monitored under future expected drier and warmer climatic scenarios and, particularly, if high management pressure remains.

@@

We found similar tree competence levels in our sampled sites (Table 1), but differences in tree size and age suggest different management origin (*i.e.* land-use legacy). Our study sites had different land-use trajectories driven by the differential pattern of the natural resource use [@JimenezOlivencia2015; @MartinCivantos2014].

In Northern sites (e.g. San Juan), there is a sequentially distribution of the land uses along the elevational gradient [CITA CATALOGO]. Grasslands and shrublands for cattle farming are located at high elevations. Then forests formation with some croplands; and then, at lowlands, irrigated terraces with tree crops. Another activities could conditioning the natural resources uses. For instance, the mining activity in San Juan site have strongly conditioned the forest structure. This oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history (citas).  Historical documents indicated two periods of intense mining activity: the second half of the 19th century after the publication of detailed mineralogical reports by [@MaestreXXXX]; and during the first decades of the twentieth century until 1960, which is the last year in which there is evidence of the existence of mining activity [cita].

On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands, even reached high elevation (mainly barley, rye and potatoes) [@Calatrava2019]. Forest resources, like firewood, charcoal, acorns, have been continuously exploited in this site through history [@ValbuenaCarabana2014; @JimenezSerrano2004]. In Cáñar there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] up to middle of the 20th century. From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment [@MesaTorres2009; @Bonet2014_conama].

The different patterns of resource uses can help to understand the tree-growth changes recorded in our chronologies (Figure 5). We observed a tree-growth release event around 1940 in the SJ oak woodland (Figures 5, S4) which concurs with one period of mining activity for this area. Several documentary sources indicate an apogee of the mining activity during 1925 to 1957 period, supported by both the improvement of the paths to transport the mineral and by the creation of new railway connections (CITA). During this period there were an increase on the use of timbers for the tunnels of the mines (CITE) and several furnaces that required great quantities of fuelwood to melt the mineral were active in this area (cite). This heavily exploitation of the forest resources could affect to a major part of this oak woodland, since percentage of trees affected by GC > 50 % reaches values above 50 % (Figure S4). Our results also coincide with a concurrent but less intense tree-growth release event reported for a closed oak woodland [@GeaIzquierdo2014]. In addition, paleoecological studies carried out in nearby alpine bogs have recorded increases in the heavy metals concentration since the end of 18th century until mid-20th, which coincides with the maximum activity of the mining in this area [@GarciaAlix2017].

@@

Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution [*e.g.* @GeaIzquierdo2014; @Herrero2013; @Matias2017], but in Mediterranean forests we must also considered the management history, *i.e* *land-use legacies* [@Penuelas2017; @DoblasMiranda2017; @Navarro2013], since, as we inferred from our results, the land-use legacy can conditioned tree-growth and resilience of tree species.

>>>>

* The studied relict populations showed expressions of management, likely reflecting the origin of the current forest structure. Both resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and also to long-term climatic changes (*i.e.* warming), despite the ‘a priori’ vulnerability that we were expecting them to express in respone climate change on relict stands resulting from intensive land-use in history at their current rear-edge. Furthermore, all stands show a positive growth trend (stronger in high elevation stand) since the end of the 1970s matching the simultaneous increase in temperature observed.

* Resilience to drought followed a dryness gradient among the three populations studied: Northern populations thriving under a drier climate were less resilient to drought both in primary and secondary growth than Southern populations where lower altitudes were also less resilient than those at high-altitudes.

* Despite being considered relict stands at the dry rear-edge for the species, they showed high resilience in response to drought.

Yet, the high sensitivity to water availability expressed and positive growth trends only in the coldest and more humid stands suggest divergent response to a changing climate along very short climatic gradients and suggest that dynamics (i.e. not only adult performance but also regeneration) of those stands located at the lowest elevations, hence the altitudinal rear-edge, need to be monitored under future expected drier and warmer climatic scenarios and, particularly, if high management pressure remains.

@@

We found similar tree competence levels in our sampled sites (Table 1), but differences in tree size and age suggest different management origin (*i.e.* land-use legacy). The review of historical documents revealed that our study sites had different land-use trajectories driven mainly by the differential pattern of the natural resource use [@JimenezOlivencia2015; @MartinCivantos2014]. On the one hand, the northern sites (*e.g.* San Juan), show a sequentially distribution of the land uses along the elevational gradient [Catalogo Paisajes]: grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops. Other activities (*e.g.* mining) could also determine the use of natural resources, and therefore the forest structure. For instance, the SJ oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history [quarries]. Historical documents indicated two periods of intense mining activity: the second half of the 19th century after the publication of detailed mineralogical reports by [@MaestreXXXX]; and during the first decades of the twentieth century until 1960, which is the last year in which there is evidence of the existence of mining activity [mineria1950]. On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands, even reached high elevation (mainly barley, rye and potatoes) [@Calatrava2019]. Forest resources, like firewood, charcoal and acorns, have been continuously exploited in southern sites through history [@ValbuenaCarabana2014; @JimenezSerrano2004]. In Cáñar site, there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] up to middle of the 20th century. From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment [@MesaTorres2009; @Bonet2014_conama].

Those different patterns of resource uses can help to understand the tree-growth changes recorded in our chronologies (Figure 5). We observed a tree-growth release event at 1940s in the SJ oak woodland (Figures 5, S4) which concurs with one period of maximum mining activity for this area. Several documentary sources indicate an apogee of the mining activity during 1925 to 1957 period, supported by both the improvement of the paths to transport the mineral and by the creation of new railway connections [tranvia]. During this period there were an increase on the use of timbers for the tunnels of the mines (CITE) and several furnaces that required great quantities of fuelwood to melt the mineral were active in this area (cite). This heavily exploitation of the forest resources could affect to a major part of this oak woodland, since percentage of trees affected by GC > 50 % reaches values above 50 % (Figure S4). Our results also coincide with a concurrent but less intense tree-growth release event reported for a closed oak woodland [@GeaIzquierdo2014]. In addition, paleoecological studies carried out in nearby alpine bogs have recorded increases in the heavy metals concentration since the end of 18th century until mid-20th, which coincides with the maximum activity of the mining in this area [@GarciaAlix2017].

@@

Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution [*e.g.* @GeaIzquierdo2014; @Herrero2013; @Matias2017], but in Mediterranean forests we must also considered the management history, *i.e* *land-use legacies* [@Penuelas2017; @DoblasMiranda2017; @Navarro2013; @Ameztegui2016] since, as we inferred from our results, the land-use legacy can conditioned tree-growth and resilience of tree species. Although the abandonment of the traditional activities have provoked a stagnation for this species [@Canellas2004], we have observed an increase in the growth, especially in the high elevation sites, where the water is less-limiting resource.

Conclussions

Severe drought events have provoked reductions in primary and secondary growth *Q. pyrenaica* populations located in the rear-edge of their distribution.

Despite the expected vulnerability for those relict stands, we observed positive trends in primary growth (since 2000) and secondary growth (since 1970) of *Q. pyrenaica* in Sierra Nevada mountain range. Surprisingly we also observed

Those trends could have relation with the

A corto plazo, la tendencia positiva observada tanto para el crecimiento primario como para el secundario podría tener relación con varios factores. Por un lado

La tendencia creciente del crecimiento en el crecimiento observada en los últimos años podría relacionarse con un abandono de las practicas tradicionales en esta region de montaña, así como 
el abandono de las prácticas tradicionales así como por el caracter de re

El abandono de las masas de Q. pyrenaica ha provocado un estado general de degradación, que puede tener consecu

La reducción de la presión antrópica durante las últimas décadas ha provocado que estas bosques presenten un avanzado estado de degradación

After at least five decades of abandonment, old Cs of Q. pyrenaica in Spain present a general state of degradation

In this work, molecular tools have been successfully applied to assess levels of genetic diversity in Cs, rejecting one of the most widespread and preconceived idea in Spanish silviculture: the decline in genetic diversity promoted by historical coppicing in Q. pyrenaica.
